---
title: "Zambia MAIZE Forescasts using MODIS VI indices and RHEAS"
author: "Dr. -Ing. MISK Benson Kenduiywo"
csl: "apa-6th-edition.csl"
output: 
   html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r, global_options, tidy = TRUE, echo=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60))
opts_knit$set(global.par = TRUE)
```


## Introduction

We will use crop yield data from the Zambian Ministry of Agriculture Livestock to develop a model to predict yields in different districts The data is available for the years 2011--2022.

Load necessary packages and define directories.

```{r refdata, echo=FALSE, warning=FALSE, message=FALSE}
rm(list = ls(all=TRUE))
unlink(".RData")
library(dplyr)
library(reshape2)
root <- "/home/servir/RHEAS/data/zambia/"
metric <- paste0(root, "metrics/")
ref <- paste0(root, "ref/")

```

## Reference data

Load the maize yield forecasts as per the Zambian Ministry of Agriculture.

```{r y2, echo=FALSE, warning=FALSE, message=FALSE}
#Load 2010-2018 data
ref <- read.csv(paste0(ref, "Zambia_District _Maize_Forecasting_2011_2022.csv"), stringsAsFactors =  FALSE)
ref$District <- toupper(ref$District)
ref$Province <- trimws(toupper(ref$Province), which = c("both"))
sort(unique(ref$Province))
ref$District <- trimws(toupper(ref$District), which = c("both"))
names(ref)[7] <- "year"
```

Visualize the crop forecasting data from MoA per year.

```{r par, echo=FALSE, warning=FALSE, message=FALSE}
par(mar=c(2,2,2,2)) #c(bottom, left, top, right)
```


```{r ref3, echo=FALSE, warning=FALSE, message=FALSE}

boxplot(yield_MT_ha~year, data=ref, col=rainbow(length(unique(ref$Year))), xlab="Year", ylab = "Yield (MT/ha)", main="Zambia MoA Annual Forecasts.")

```

Visualize the crop forecasting data from MoA per Province.

```{r ref4, echo=FALSE, warning=FALSE, message=FALSE}
agg <- ref
agg$Province <- substr(agg$Province,start=1,stop=4)
agg <- na.omit(agg)
boxplot(yield_MT_ha~Province, data=agg, col=rainbow(length(unique(agg$Province))), xlab="Provinces", ylab = "Yield (MT/ha)", main="Zambia MoA Provinces Forecasts.", cex.axis=0.8) #horizontal = T, las=1

```


## MODIS RS Indices

### Get Indices

Load MODIS Normalized Difference Vegetation Index (NDVI), Green Normalized difference Vegetation Index (GNDVI), Normalized Difference Moisture Index (NDMI), Gross Primary Productivity (GPP) and Fraction of Photosynthetically Active Radiation (FPAR) county aggregates by date for the period 2000--2017 and convert to long format.

```{r modis, echo=FALSE, warning=FALSE, message=FALSE}
wide2long <- function(df, variable){
  melt(df[, -c(1,45)], variable.name = "date", 
          value.name = variable, id.vars = "county")
}
#Indices list; https://www.l3harrisgeospatial.com/docs/canopywatercontent.html#:~:text=Moisture%20Stress%20Index%20(MSI),absorption%20around%201599%20nm%20increases.
#1.0 EVI
temp <- list.files(metric, pattern=glob2rx("*_evi_*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
evi <- lapply(files, wide2long, "evi")
evi[[1]][1,]

#2.0 FPAR
temp <- list.files(metric, pattern=glob2rx("*fpar*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
fpar <- lapply(files, wide2long, "fpar")
fpar[[1]][1,]

#3.0 GLI
temp <- list.files(metric, pattern=glob2rx("*_gli_*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
gli <- lapply(files, wide2long, "gli")
gli[[1]][1,]

#4.0 GNDVI
temp <- list.files(metric, pattern=glob2rx("*gndvi*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
gndvi <- lapply(files, wide2long, "gndvi")
gndvi[[1]][1,]

#5.0 GPP
temp <- list.files(metric, pattern=glob2rx("*gpp*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
gpp <- lapply(files, wide2long, "gpp")
gpp[[1]][1,]

#6.0 LAI
temp <- list.files(metric, pattern=glob2rx("*lai*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
lai <- lapply(files, wide2long, "lai")
lai[[1]][1,]

#7.0 MSI
temp <- list.files(metric, pattern=glob2rx("*_msi_*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
msi <- lapply(files, wide2long, "msi")
msi[[1]][1,]

#8.0 NDMI (Normalized Difference Moisture Index)
temp <- list.files(metric, pattern=glob2rx("*ndmi*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
ndmi <- lapply(files, wide2long, "ndmi")
ndmi[[1]][1,]

#9.0 NDVI
temp <- list.files(metric, pattern=glob2rx("*_ndvi_*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
ndvi <- lapply(files, wide2long, "ndvi")
ndvi[[1]][1,]

#10.0 NPCRI
temp <- list.files(metric, pattern=glob2rx("*_npcri_*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
npcri <- lapply(files, wide2long, "npcri")
npcri[[1]][1,]

#11.0 SIPI
temp <- list.files(metric, pattern=glob2rx("*_sipi_*"))
files <-lapply(paste0(metric,temp), read.csv, stringsAsFactors =  FALSE)
sipi <- lapply(files, wide2long, "sipi")
sipi[[1]][1,]

```

Format the date variable by removing unnecessary characters.

```{r y1, echo=FALSE, warning=FALSE, message=FALSE}
formatDate <- function(m){
  m$date <- gsub("X","", m$date)
  m$date <- gsub("_nd","", m$date)
  m$date <- as.Date(as.character(m$date), format = "%Y_%m_%d")
  m[1,]
  colnames(m)[1]<- "District"
  return(m)
}
evi <- lapply(evi, formatDate)
evi[[1]][1,]
fpar <- lapply(fpar, formatDate)
fpar[[1]][1,]
gli <- lapply(gli, formatDate)
gli[[1]][1,]
gndvi <- lapply(gndvi, formatDate)
gndvi[[1]][1,]
gpp <- lapply(gpp, formatDate)
gpp[[1]][1,]
lai <- lapply(lai, formatDate)
lai[[1]][1,]
msi <- lapply(msi, formatDate)
msi[[1]][1,]
ndmi <- lapply(ndmi, formatDate)
ndmi[[1]][1,]
ndvi <- lapply(ndvi, formatDate)
ndvi[[1]][1,]
npcri <- lapply(npcri, formatDate)
npcri[[1]][1,]
sipi  <- lapply(sipi, formatDate)
sipi[[1]][1,]

```

Now combine the list into one dataframe.

```{r y1b, echo=FALSE, warning=FALSE, message=FALSE}
evi <- do.call(rbind.data.frame, evi)
fpar <- do.call(rbind.data.frame, fpar)
gli  <- do.call(rbind.data.frame, gli)
gndvi  <- do.call(rbind.data.frame, gndvi)
gpp <- do.call(rbind.data.frame, gpp)
lai <- do.call(rbind.data.frame, lai)
msi <- do.call(rbind.data.frame, msi)
ndmi <- do.call(rbind.data.frame, ndmi)
ndvi <- do.call(rbind.data.frame, ndvi)
npcri <- do.call(rbind.data.frame, npcri)
sipi <- do.call(rbind.data.frame, sipi)

```


Now merge all the remote sensing metrics to one data frame.

```{r y1c, echo=FALSE, warning=FALSE, message=FALSE}
evi$evi <- as.numeric(evi$evi)
n <- do.call("cbind", list(evi, gli=as.numeric(gli[,-c(1:2)]), gndvi=as.numeric(gndvi[,-c(1:2)]), gpp=as.numeric(gpp[,-c(1:2)]), msi=as.numeric(msi[,-c(1:2)]), ndmi=as.numeric(ndmi[,-c(1:2)]), ndvi=as.numeric(ndvi[,-c(1:2)]), npcri=as.numeric(npcri[,-c(1:2)]), sipi=as.numeric(sipi[,-c(1:2)])))
temp  <- cbind(lai, fpar=fpar[,-c(1:2)])
temp[, c("lai","fpar")] <- lapply(c("lai","fpar"), function(x) as.numeric(temp[[x]]))
modis <- merge(n,temp, by=c("District", "date"))
modis$District[modis$District=="CHIENGI"] <- "CHIENGE"
modis$District[modis$District=="SHANGOMBO"] <- "SHANG'OMBO"
```


### RS Temporal Index Aggregation 

Aggregate metrics with respect to District maize growing calendar based on only the vegetative. There is only one season in Zambia.
Define a function to capture the seasons of Maize growth in Zambia.

```{r st1, echo=FALSE, warning=FALSE, message=FALSE}
seasonMean <- function(year, df, seasons=1:2) {
  res <- list()
  for (i in seasons) {
    season <- ifelse(i==1, "long", "short")
    if (season =="long") {
      sdate <- paste0(year, "-03-01")
      edate <- paste0(year, "-09-30")
      season <- "LRLD"
    } else if (season =="short") {
      sdate <- paste0(year-1, "-10-01")
      edate <- paste0(year, "-06-30")
      season <- "SRSD"
    } else {
      stop("Define season")
    }
    ydf <- df[df$date >= sdate & df$date <= edate, ]
    #ym <- aggregate(ydf[,3], ydf[,1, drop=FALSE], mean, na.rm=T)
    ydf <- subset(ydf, select=-date)
    ym <- aggregate(.~District, data=ydf, mean, na.rm=T)
    ym$year <- year
    ym$season <- season
    res[[i]] <- ym    
  }  
  do.call(rbind, res)  
}

```

Temporally aggregate the spatial indices to obtain *spatial-temporal RS metrics*.

```{r st2, echo=FALSE, warning=FALSE, message=FALSE}
years <- 2010:2022
temp <- lapply(years, seasonMean, modis, seasons=2)
rs <- do.call(rbind, temp)
rs <- rs[!rs$District=="Counties",]
```

## DSSAT Indices

Load and aggregate RHEAS simulated Leaf Area Index (LAI), Water stress and Grain Weight Average Dry (GWAD) across different ensembles. Extract year from dates (we will use harvest year).


```{r dssat1, echo=FALSE, warning=FALSE, message=FALSE}
path <- "/home/servir/RHEAS/yields/"
tt <- read.csv(paste0(path, "zambia_chirps_dssatTable_1990_2022.csv"), stringsAsFactors =  FALSE)
tt$harvest <- as.Date(tt$harvest)
tt$planting <- as.Date(tt$planting)
tt$date <- format(tt$harvest, format = "%Y")
names(tt)[3] <- "District"
tt$District[tt$District=="CHIENGI"] <- "CHIENGE"
tt$District[tt$District=="SHANGOMBO"] <- "SHANG'OMBO"
```


### DSSAT Spatial-Temporal metrics

Aggregate DSSAT production forecasts and metrics with respect to Districts and maize growing calendar. 

The maize growing season in Zambia starts from October to end of June. So we will aggregate the metrics and forecast with this condition using the function `RH_metrics`.

```{r rh2, echo=FALSE, warning=FALSE, message=FALSE}
RH_metrics <- function(rh, sStart, sEnd){
  rh <- subset(rh, format(as.Date(rh$planting), "%m") >= sStart & format(as.Date(rh$harvest), "%m") <= sEnd)
  rh <- aggregate(rh[,c("wsgd","lai","gwad")], rh[,c("District","date")], mean, na.rm=T)
  return(rh)
}

rh <- RH_metrics(tt, sStart ="10", sEnd = "06")
#rh <- subset(rh, select = - gwad)
names(rh)[2] <- "year"

```

Check and format District names to be consistent in all datasets.

```{r naming, echo=FALSE, warning=FALSE, message=FALSE}
c <- sort(unique(ref$District))
c[!c %in% sort(unique(rs$District))]
c[!c %in% sort(unique(rh$District))]
```


### Add VIC Spatial-Temporal metrics

Load spatial-temporal metrics for VIC model.

```{r vic1, echo=FALSE, warning=FALSE, message=FALSE}
vc <- readRDS(paste0(metric,"vic_metrics_v1.rds"))
#vc <- subset(vc, select=-c(season, prec, rainf, evap_canop))
vc <- subset(vc, select=-c(season))
```


## Feature Engineering

Here we explore how different features perform w.r.t yield forecasting.

Normalize all indices to between 0 and 1.

```{r feat1, echo=FALSE, warning=FALSE, message=FALSE}
minMax <- function(x){
  return((x-min(x, na.rm=T))/(max(x, na.rm = T) - min(x, na.rm=T)))
}
```


Combine MODIS NDVI with reference data and including RHEAS indices as a separate dataframe for comparison.

```{r feat2, echo=FALSE, warning=FALSE, message=FALSE}
rs <- subset(rs, select=-season)
rs[,-c(1,13)] <- apply(rs[,-c(1,13)], 2, minMax)
vc[,-c(1,13)] <- apply(vc[,-c(1,13)], 2, minMax)
names(rh)[4] <- "DSSAT_lai" 
rh[,-c(1,2)] <- apply(rh[,-c(1,2)], 2, minMax)

df_list <- list(rs[rs$year > 2010,], rh[rh$year > 2010,], vc[vc$year > 2010, ], ref[,c("District", "yield_MT_ha","year")])
data <- Reduce(function(x, y) merge(x, y, by=c("District","year")), df_list)
df_list2 <- list(rs[rs$year > 2010,], ref[,c("District", "yield_MT_ha","year")])
data <- subset(data, select=-gwad)
vi <- Reduce(function(x, y) merge(x, y, by=c("District","year")), df_list2)
  
```


Feature selection using random forest

```{r feat3, echo=FALSE, warning=FALSE, message=FALSE}
library(randomForest)
library(ggplot2)
library(ggthemes)
library(dplyr)
rf = randomForest(yield_MT_ha~., data=subset(data, select = -c(District,year)), importance=TRUE, ntree = 500)
importance <- importance(rf)
importance
rf
varImportance <- data.frame(Variables = row.names(importance), 
                            Importance = round(importance[ ,'%IncMSE'],2))

#Create a rank variable based on importance
rankImportance <- varImportance %>%
  mutate(Rank = paste0('#',dense_rank(desc(Importance))))

#Use ggplot2 to visualize the relative importance of variables

ggplot(rankImportance, aes(x = reorder(Variables, Importance), 
                           y = Importance, fill = Importance)) +
  geom_bar(stat='identity') + 
  geom_text(aes(x = Variables, y = 0.5, label = Rank),
            hjust=0, vjust=0.55, size = 4, colour = 'red') +
  labs(x = 'EO Metrics') +
  coord_flip() + 
  theme_few(base_size = 14)

## Select features that have an impact of 10% MSE on prediction
#selected <- rankImportance$Variables[rankImportance$Importance>=12]
#data <- subset(data, select=selected)
```


## Validation

### RMSE

We can use the Root Mean Square Error (RMSE) to evaluate the models accuracy. RMSE is given as:

$$
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i=1}^n \widehat{y}-y},
$$
where $\widehat{y}$  and $y$ are predicted yields and observed yields respectively while *n* is the number of fitted points.

```{r rmse, echo=FALSE, warning=FALSE, message=FALSE}
rmse <- function(error){
  sqrt(mean(error^2, na.rm=T))
}

```

### MAPE

Mean absolute percentage error (MAPE) is given as:

$$
\text{MAPE} = \frac{100\%}{n} \sum_{i=1}^n |\frac{y-\widehat{y}}{y}|.
$$
In R we can write it as:

```{r mape, echo=FALSE, warning=FALSE, message=FALSE}
MAPE <- function (y_pred, y_true){
    MAPE <- mean(abs((y_true - y_pred)/y_true))
    return(MAPE*100)
}
```

### $R^2$

We can also compute $R^2$ measure as follows:

```{r r_Squared, echo=FALSE, warning=FALSE, message=FALSE}
R_square <- function(actual, predicted) {
  val <- 1 - (sum((actual-predicted)^2)/sum((actual-mean(actual))^2))
  val
} 
```

### RRMSE

What about Relative root mean square error (RRMSE):

$$
\text{RRMSE} = \frac{\sqrt{\frac{1}{n} \sum_{i=1}^n y-\widehat{y}}}{\frac{1}{n} \sum_{i=1}^n y},
$$
According to [Li et al. 2013](ttp://dx.doi.org/10.1016/j.enconman.2013.03.004), the performance of the model is excellent when RRMSE < 10%; good if 10\% < RRMSE < 20\%; fair if 20\% < RRMSE < 30\%; and poor if RRMSE ≥ 30\%.

```{r rrmse, echo=FALSE, warning=FALSE, message=FALSE}
rrmse <- function(predicted, observed){
  error <- observed - predicted
  re <- sqrt(mean(error^2))/mean(observed)
  return(re*100)
}
```

CNN Model

```{r cnn}
library(keras)
library(caret)

set.seed(123)
cnn <-  function(xtrain, ytrain, in_dim){
	model = keras_model_sequential() %>%
	layer_conv_1d(filters = 64, kernel_size = 2,
                input_shape = in_dim, activation = "relu") %>%
	layer_flatten() %>%
	layer_dense(units = 32, activation = "relu") %>%
	layer_dense(units = 1, activation = "linear")

	model %>% compile(
	  loss = "mse",
	  optimizer = "adam")
	model %>% summary()
	model %>% fit(xtrain, ytrain, epochs = 100, batch_size=16, verbose = 0)
	scores = model %>% evaluate(xtrain, ytrain, verbose = 0)
	print(scores)
	return(model)
}

```

Leave one year out cross validation using RS metrics only. Createa modelling function comparing RF and SVM.

```{r cv1, echo=FALSE, warning=FALSE, message=FALSE}
#library(rJava)
#Sys.setenv(JAVA_HOME="C:/Program Files/Java/jdk-18.0.1.1")
#options(java.parameters = "-Xmx4g")
#extrees <- "https://cran.r-project.org/src/contrib/Archive/extraTrees/extraTrees_1.0.5.tar.gz"
#install.packages(extrees, repos=NULL, type="source", dependencies = T) 
#rjv <- "https://cran.r-project.org/src/contrib/Archive/rJava/rJava_1.0-5.tar.gz"
#install.packages(rjv, repos=NULL, type="source", dependencies = T)

#library(extraTrees)
library(dismo)
library(e1071)

models <- function(vi, years, accName){
  npredictors <- dim(vi)[2]
  svm_a <- c()
  svm_b <- c()
  rf_a <- c()
  rf_b <- c()
  r_svm <- c()
  r_rf <- c()
  cnn_r <- c()
  cnn_rmse <- c()
  cnn_mape <- c()
  df <- na.omit(subset(vi, select=-District))
  #df$District <- as.factor(df$District)
  y <- years
  for(i in 1:length(y)){
    observed_y <- 0
    svm_y <- 0
    print(paste0('The year ', y[i], " left out for validation.\n"))
    train <- subset(df, year != y[i], select=-year)
  	valid <- subset(df, year == y[i], select=-year)
  	observed_y <- valid$yield_MT_ha
    #SVM
    tuneResult <- tune(method="svm", yield_MT_ha~.,  data = train, ranges = list(epsilon = seq(0,1,0.1), cost = (seq(0.5,8,.5))), kernel="radial" )
    #svm <- svm(Yield_MT_HA~., data=data[, c("Yield_MT_HA","gndvi", "ndvi","ndmi", "gpp", "fpar", "Region")], kernel="radial" , cross=5)
    svm_y <- predict(tuneResult$best.model, valid)
    #svm_y <- predict(svm, valid)
    svm_a[i] <- rmse(observed_y-svm_y)
    svm_b[i] <- MAPE(observed_y, svm_y)
    cat("SVM Coefficient of determination R^2\n")
    r_svm[i] <- R_square(observed_y, svm_y)
    print(r_svm)
    #RF
    tuneRF <- tune(method="randomForest", yield_MT_ha~.,  data = train, ranges = list(ntree = c(100, 500), mtry = seq(1,npredictors,1))) #
    print(tuneRF$best.model)
    #rf = randomForest(Yield_MT_HA~., data=train, importance=TRUE, ntree = 500)
  
    rf_y <- predict(tuneRF$best.model, valid)
    rf_a[i] <- rmse(observed_y-rf_y)
    rf_b[i] <- MAPE(observed_y, rf_y)
    cat("RF Coefficient of determination R^2\n")
    r_rf[i] <- R_square(observed_y, rf_y)
    print(r_rf)
    
    # #CNN
    # y <-  as.matrix(train[ , "yield_MT_ha"])
    # x <- array(unlist(subset(train, select=-yield_MT_ha)), dim = c(nrow(train), ncol(train), 1))
    # xtest <- array(unlist(valid), dim = c(nrow(valid), ncol(valid), 1))
    # in_dim <- c(dim(x)[2:3])
    # cnns <- cnn(x, y, in_dim)
    # cnn_y <- predict(cnns, xtest)
    # cnn_rmse[i] <- rmse(observed_y-cnn_y)
    # cnn_mape[i] <- rrmse(observed_y, cnn_y)
    # cat("CNN Coefficient of determination R^2\n")
    # cnn_r[i] <- R_square(observed_y, cnn_y)
    # print(cnn_r)
    
  }
  
  cat("SVM model RMSE is ", mean(svm_a), "\n")
  cat("SVM model MAPE is ", mean(svm_b), "\n")
  cat("SVM model R2 is ", mean(r_svm), "\n")
  cat("RF model RMSE is ", mean(rf_a), "\n")
  cat("RF model R2 is ", mean(r_rf), "\n")
  cat("RF model MAPE is ", mean(rf_b), "\n")
  #cat("CNN model RMSE is ", mean(cnn_rmse), "\n")
  #cat("CNN model R2 is ", mean(cnn_r), "\n")
  #cat("CNN model MAPE is ", mean(cnn_mape), "\n")
  
  temp <- rbind(data.frame(RMSE=rf_a, Method="RF", Year=y), data.frame(RMSE=svm_a, Method="SVM", Year=y))
  #temp <- rbind(temp, data.frame(RMSE=cnn_rmse, Method="CNN", Year=y))

  par(mfrow=c(2,2), mar=c(4.5,4.5,1,1))
  boxplot(RMSE ~ Method, data =temp, col=c("#999999", "#E69F00"), ylab="RMSE (tons/ha)", xlab="")
  acc <- temp
  
  temp <- rbind(data.frame(MAPE=rf_b, Method="RF", Year=y), data.frame(MAPE=svm_b, Method="SVM", Year=y))
  #temp <- rbind(temp, data.frame(MAPE=cnn_mape, Method="CNN", Year=y))
    
   boxplot(MAPE ~ Method, data =temp, col=c("#999999", "#E69F00"), ylab="MAPE (%)", xlab="", ylim= c(0,100))
  acc <- merge(acc, temp, by=c("Method", "Year"))
   
  temp <- rbind(data.frame(R2=r_rf, Method="RF", Year=y), data.frame(R2=r_svm, Method="SVM", Year=y))
  #temp <- rbind(temp, data.frame(R2=cnn_r, Method="CNN", Year=y))
  
  boxplot(R2 ~ Method, data =temp, col=c("#999999", "#E69F00"), ylab=expression(R^2), xlab="", ylim= c(0,1))
   
   acc <- merge(acc,temp, by=c("Method", "Year"))
   fileName <- paste0(accName,".rds")
   saveRDS(acc, fileName)
}

models(vi, years = 2011:2022, accName = "EO_only")

```

Now include RHEAS metrics and see how accuracy behaves.

```{r cv2, echo=FALSE, warning=FALSE, message=FALSE}
models(data, years = 2011:2022, accName = "EO_RHEAS")
```

## Predict current Year Yields

Use best configuration to predict current season yields.




