---
title: "Tanzania Maize Yield Forecast"
author: "Dr. -Ing. MISK Benson Kenduiywo"
csl: "apa-6th-edition.csl"
output: 
   html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60))
opts_knit$set(global.par = TRUE)
```

## RHEAS forecasts

Load and aggregate RHEAS simulated Leaf Area Index (LAI), Water stress and Grain Weight Average Dry (GWAD) across different ensembles. Extract year from dates (we will use harvest year).

```{r rh1, echo=FALSE, message=F}
rm(list = ls(all=TRUE))
unlink(".RData")
root <- "D:/RCMRD/Data/Yields/"

formatdf <- function(filename){
  tt <- read.csv(filename, stringsAsFactors =  FALSE)
  tt$harvest <- as.Date(tt$harvest)
  tt$planting <- as.Date(tt$planting)
  tt$date <- format(tt$harvest, format = "%Y")
  names(tt)[3] <- "Region"
  tt$Region <- toupper(tt$Region)
  return(tt)
}

bi <- formatdf(paste0(root, "RHEAS/tanzania_tamsat_25km_districts_bimod_dssatTable_2000_2022_100kg.csv"))

un <- formatdf(paste0(root, "RHEAS/tanzania_tamsat_25km_districts_unimod_dssatTable_2000_2022_100kg.csv"))
```


## Metrics aggregagation

Aggregate RHEAS production forecasts and metrics with respect to Districts maize growing calendar. 

We assume three maize growing seasons in Tanzania:

1) Season 1: OND (SRSD) October -- February/March.
2) Season 2: MAM March -- September. 
3) Season 3: Long Rain (LR) November -- April/August 

So we will aggregate the metrics and yield forecast per district with this condition using the function `RH_metrics`.

```{r rh2, echo=F, message=F}
RH_metrics <- function(rh, sStart, sEnd, season){
  rh <- subset(rh, format(as.Date(rh$planting), "%m") %in% sStart & format(as.Date(rh$harvest), "%m") %in% sEnd)
  rh$Season <- season
  rh <- aggregate(rh[,c("wsgd","lai","gwad"), drop=FALSE], rh[,c("Region","date", "Season"), drop=FALSE], mean, na.rm=T)
  
  
  return(rh)
}
# Unimodal season:  October--August (LR)
sStart <- c("09", "10", "11", "12", "01","02", "03", "04")
sEnd <- c("02", "03", "04", "05", "06", "07", "08")
rh_u <- RH_metrics(un, sStart, sEnd, "LR")

#Bimodal season A (OND): September--March
sStart <- c("09", "10", "11")
sEnd <- c("01","02", "03")
rh_a <- RH_metrics(bi, sStart=sStart, sEnd=sEnd, "OND")

#Bimodal season B (MAM): January--August
sStart <- c("01", "02", "03", "04")
sEnd <- c("05", "06", "07", "08")
rh_b <- RH_metrics(bi, sStart=sStart, sEnd=sEnd, "MAM")

# Long Rain
#lr <- aggregate(tt[,c("wsgd","lai","gwad"), drop=FALSE], tt[,c("Region","date"), drop=FALSE], mean, na.rm=T)
#Merge all
rh <- do.call(rbind, list(rh_u, rh_a, rh_b))

```

Convert RHEAS yields from kg/ha to MT/ha.

```{r rh3, echo=FALSE}
rh$gwad <- rh$gwad/1000
#Remove values along the lake region
rh$gwad[rh$Region=="LAKE VICTORIA"] <- NA
```


## Visualization

Add shapefile for visualization.

```{r, v1, echo=F, message=F}
library(raster)
filename <- "D:/Adm data/Tanzania/gadm36_TZA_2.shp"
tza <- shapefile(filename)
names(tza)[7] <- "Region"
tza$Region <- toupper(tza$Region)
```


Check and format Region names to be consistent in both the RHEAS and administrative boundaries. Here we check which regions are mission from RHEAS predictions.

```{r v2, echo=F, message=F}
c <- sort(unique(tza$Region))
c[!c %in% sort(unique(rh$Region))]
```

 Merge RHEAS and Admin data.
 
```{r v3, echo=F, message=F}
year <- 2022
temp <- rh[, c("Region", "date", "gwad", "lai", "wsgd", "Season")]
temp <- reshape(temp, idvar = c("Region","date"), timevar = "Season", v.names  = c("gwad", "lai", "wsgd"),direction = "wide")
tza <-  merge(tza[,c("Region", 'NAME_1')], temp[temp$date==year,], by = "Region", duplicateGeoms=TRUE)
#path <- "D:/RCMRD/Code/yieldprediction"
#shapefile(tza[,c("Region", "wsgd", "lai","gwad", "date","Season")], paste0(path,"/tanzania/tanzania_Maize_Forecasts.shp"),overwrite=T)

```


### Map of yields

Visualize RHEAS predicted yields spatially for *all seasons*. 

```{r v4, echo=F, message=F}
library(tmap)
library(mapview)
tmap_mode("view")
 map_a <- tm_shape(tza, name="Average yield (MT/ha) per Region") +
  tm_fill(col=c("gwad.OND", "gwad.MAM", "gwad.LR", "lai.LR"),   title=c("OND Yield (MT/ha)", "MAM Yield (MT/ha)", "LR Yield (MT/ha)", "LR LAI"), textNA = "No data", breaks = seq(0.3,4.5,1)) +
   tm_facets(sync = TRUE, ncol = 2, nrow=2) +
#   #tm_text("District", size = 0.75)+
   tm_layout(title.size =5, panel.label.size=6)+
   tm_format("World")
 map_a

#lf <- tmap_leaflet(map_a)
#mapshot(map_a, file = paste0(getwd(), "/TZ_Yields_2022.png"), remove_controls = c("homeButton", "layersControl"))
tmap_save(map_a, scale =1.6, dpi= 300, filename="TZ_Yields_2022.png")
##Other plotting options see http://www.nickeubank.com/wp-content/uploads/2015/10/RGIS3_MakingMaps_part1_mappingVectorData.html#multiple-layers-with-spplot
```

Analysis of the *LR*.

```{r lr1}
map_lr <- tm_shape(tza, name="Average yield (MT/ha) per Region") +
  tm_fill(col=c("gwad.LR", "lai.LR", "wsgd.LR"),   title=c("LR Yield (MT/ha)", "LR LAI", "LR water stress", "LR LAI"), textNA = "No data", breaks=c(-Inf, seq(0, 5, by=1), Inf),style="fixed") +
   tm_facets(sync = TRUE, ncol = 2, nrow=2) +
#   #tm_text("District", size = 0.75)+
   tm_layout(title.size =5, panel.label.size=6)+
   tm_format("World")
map_lr
tmap_save(map_lr, scale =1.6, dpi= 300, filename="TZ_unimodal_Yields_2022.png")

p1 <- tm_shape(tza, name="Average yield (MT/ha) per District") +
  tm_fill("gwad.LR", title="LR Yield (MT/ha)", textNA = "No data", breaks=c(-Inf, seq(1, 4, by=1), Inf),style="quantile") +
   tm_layout(title.size =5, panel.label.size=6)+
   tm_format("World")

p2 <- tm_shape(tza, name="Average LAI per District") +
  tm_fill("lai.LR", title="LR LAI", textNA = "No data", breaks=c(-Inf, seq(1, 2, by=0.25), Inf),style="quantile") +
   tm_layout(title.size =5, panel.label.size=6)+
   tm_format("World")


p3 <- tm_shape(tza, name="Average water stress per District") +
  tm_fill("wsgd.LR", title="LR water stress", textNA = "No data", breaks=c(-Inf, seq(0, 1, by=0.25), Inf),style="quantile") +
   tm_layout(title.size =5, panel.label.size=6)+
   tm_format("World")

#current.mode <- tmap_mode("plot")
plots <- tmap_arrange(p1, p2, p3, ncol =2)
plots
#tmap_mode(current.mode)

```

### Annual Seasonal Trends

Visualize maize yield trends for the last 15 years.
 
```{r bb1, echo=FALSE}
library(ggplot2)
temp <- rh[rh$date > 2007,]
p <- ggplot(temp, aes(x = date, y = gwad))
 
p <- p + geom_boxplot(
   aes(fill = Season),
   position = position_dodge(0.9) 
   ) +
   scale_fill_manual(values = c("#999999", "#E69F00", "green"))+
   ggtitle("Annual Seasonal  Maize Yield Trend")+
   labs(y = "Yield (MT/ha)", x = "Year")
 p

```
 
 
Visualize yields per region between 2000-2022 for OND season only.
 
```{r bb2, message=F, echo=F}
#temp <- rh[rh$Season=="OND",]
# par(mai=c(1,2.3,1,1))
# boxplot(temp$gwad ~ temp$Region, las=1, cex.axis=.75, horizontal=TRUE, xlab="Yield (MT/ha)", ylab="", cex=.5, col=rainbow(length(unique(temp$Region))))
```
 
 
## Total production

Total production (MT) can be estimated by multiplying unit production estimates (MT/ha) from RHEAS by area under maize in each district. 
Statistics for area under maize in different districts in Tanzania collected through National Bureau  of Statistics agricultural census survey of 2019/2020 is adopted.
 
```{r p1, message=F, echo=F, warning=F}
library(readstata13)
pa <- read.dta13(paste0(root, "Reference/Tanzania/NBS_Maize_2019_20 Agric Census/Maize_Shorty_Rainy_Season.dta"), convert.factors=T, generate.factors=T)
pb <- read.dta13(paste0(root, "Reference/Tanzania/NBS_Maize_2019_20 Agric Census/Maize_Long_Rainy_Season.dta"), convert.factors=T, generate.factors=T)
```

