---
title: "RHEAS Cultivar Calibration: Malawi"
author: "Dr. -Ing. MISK Benson Kenduiywo"
date: '2022-09-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Several cultivars from literature within Zambia and Malawi were tested and compared to observed production information in a bid to select a dominant variety for maize yield prediction.


## Reference data

Load malawi's ministry of agriculture maize forecasts data as reference.

```{r ref1, echo=FALSE, warning=FALSE, message=FALSE}
rm(list = ls(all=TRUE))
unlink(".RData")
root <- "D:/RCMRD/Data/Yields/"
r.file <- paste0(root, "Reference/Malawi/MOA/MALAWI_MOA_MAIZE_Production data.csv")
```

Load cultivars data. Basically, 12 maize cultivars were tested between 2010--2022 to 

```{r ref40, echo=FALSE, warning=FALSE, message=FALSE}
ref <- read.csv(r.file, stringsAsFactors =  FALSE)
names(ref)
```

Convert District and province names to upper case.

```{r ref5, echo=FALSE, warning=FALSE, message=FALSE}
ref$Division <- trimws(toupper(ref$Division), which = c("both"))
sort(unique(ref$Province))
ref$District <- trimws(toupper(ref$District), which = c("both"))
ref$yield_MT_ha <- as.numeric(ref$yield_MT_ha)
ref <- na.omit(ref)
knitr::kable(head(sort(unique(ref$District))), caption="Districts")
```


## Cultivars production

Load information on different cultivars production potential as predicted by RHEA.

```{r c1}
files <- list.files(paste0(root, "Cultivars_Selection/malawi/"), full.names = T)
f <- read.csv(files[1], stringsAsFactors =  FALSE)
df_list <- list()
for(i in 1:length(files)){
  f <- read.csv(files[i], stringsAsFactors =  FALSE)
  f$cultivar <- sub("\\_.*", "", basename(files[i]))
  temp <- sub("_v2", "", tools::file_path_sans_ext(basename(files[i])))
  f$Year <- sub(".*\\_|.csv.*", "", temp)
  df_list[[i]] <- f
}

df <- do.call(rbind, df_list)
df$cname <- toupper(df$cname)
df$gwad <- df$gwad/1000
```

Check missing names.

```{r c2}
c <- sort(unique(ref$District))
c[!c %in% sort(unique(df$cname))]
```


Correct spellings 

```{r sp}
df$cname[df$cname=="DOWA EAST"] <- "DOWA EAST"
df$cname[df$cname=="DOWAEAST"]  <- "DOWA EAST"
df$cname[df$cname=="DOWAEST"]   <- "DOWA EAST"
c <- sort(unique(ref$District))
c[!c %in% sort(unique(df$cname))]
```

Merge RHEAS and observed MOA data

```{r c3}
rh_q <- aggregate(gwad~cname+cultivar+Year, data=df, FUN = 'quantile', probs=c(0.25, 0.5, .75, .95), na.rm=T)
rh_q <- do.call(data.frame, rh_q)
names(rh_q) <- gsub("\\.","",names(rh_q))
rh_m <-aggregate(gwad~cname+cultivar+Year, data=df, FUN = mean, na.rm=T) #aggregate(df[,"gwad"], df[,c("cname","cultivar", "Year")], mean, na.rm=T)
names(rh_m)[4] <- "gwadmean"
rh <- rh_q
rh$gwadmean <- rh_m$gwadmean
names(rh)[1] <- "District"
ref.agg <-aggregate(yield_MT_ha~District+Year, data=ref, FUN = mean, na.rm=T)
dff <- merge(ref.agg,rh, by=c("District", "Year"))
```


## Validation

### RMSE
We can use the Root Mean Square Error (RMSE)  and mean absolute percentage error (MAPE) to evaluate the models accuracy. RMSE is given as:

$$
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i=1}^n y-\widehat{y}},
$$
where $\widehat{y}$  and $y$ are predicted yields and observed yields respectively while *n* is the number of fitted points.

```{r rmse, echo=FALSE, warning=FALSE, message=FALSE}
rmse <- function(error){
  sqrt(mean(error^2, na.rm=T))
}

```


### RRMSE

What about Relative root mean square error (RRMSE):

$$
\text{RRMSE} = \frac{\sqrt{\frac{1}{n} \sum_{i=1}^n y-\widehat{y}}}{\frac{1}{n} \sum_{i=1}^n y},
$$


```{r rrmse, echo=FALSE, warning=FALSE, message=FALSE}
rrmse <- function(predicted, observed){
  error <- observed - predicted
  re <- sqrt(mean(error^2, na.rm=T))/mean(observed, na.rm=T)
  return(re*100)
}
```


According to [Li et al. 2013](ttp://dx.doi.org/10.1016/j.enconman.2013.03.004), the performance of the model is excellent when RRMSE < 10%; good if 10\% < RRMSE < 20\%; fair if 20\% < RRMSE < 30\%; and poor if RRMSE ≥ 30\%.


### Compute validation measures

Here validation measures are computed for each cultivar at district level based on historically observed maize yields. 

```{r v1}
val <- data.frame()
Cultivar <- sort(unique(dff$cultivar))
dists <- sort(unique(dff$District))
for(i in 1:length(Cultivar)){
  for(j in 1:length(dists)){
    #print(paste0("cultivar",i, " in ", dists[j]))
    temp <- subset(dff, cultivar==paste0("cultivar",i) & District==dists[j])
    rmse_m <- rmse(temp$gwadmean-temp$yield_MT_ha)
    rrmse_m <- rrmse(temp$gwadmean, temp$yield_MT_ha)
    pcor_m <- cor(temp$gwadmean, temp$yield_MT_ha, method = "pearson")
    bias <- mean(temp$gwadmean-temp$yield_MT_ha)
    rmse25 <- rmse(temp$gwad25-temp$yield_MT_ha)
    rrmse25 <- rrmse(temp$gwad25, temp$yield_MT_ha)
    rmse50 <- rmse(temp$gwad50-temp$yield_MT_ha)
    rrmse50 <- rrmse(temp$gwad50, temp$yield_MT_ha)
    rmse75 <- rmse(temp$gwad75-temp$yield_MT_ha)
    rrmse75 <- rrmse(temp$gwad75, temp$yield_MT_ha)
    rmse95 <- rmse(temp$gwad95-temp$yield_MT_ha)
    rrmse95 <- rrmse(temp$gwad95, temp$yield_MT_ha)
    output <- c(i, dists[j], rmse_m, rrmse_m, pcor_m, rmse25, bias, rrmse25, rmse50, rrmse50, rmse75, rrmse75, rmse95, rrmse95)
    val <-  rbind(val, output)
  }
  
}


colnames(val) <-c("Cultivar", "District", "RMSE_mean", "RRMSE_mean", "pcor", "bias", "RMSE25", "RRMSE25", "RMSE50", "RRMSE50", "RMSE75", "RRMSE75", "RMSE95", "RRMSE95")

#knitr::kable(val,caption = "Cultivar assessement based on ensemble mean, 25%. 50%, 75% and 95% percentiles respectively.")
```

Find cultivars with least RMSE in each district.

```{r v2}
vals <- data.frame()
for(i in 1:length(dists)){
  #Flter out district
  temp <- subset(val, District==dists[i])
  #Order dataframe in decreasing correlation value
  index <- order(temp$pcor, decreasing = T)
  #pick dataframe with first four highest correlation
  temp <- temp[index[1:4],]
  output <- c(temp$Cultivar[which.min(temp$RMSE_mean)], dists[i], min(temp$RMSE_mean))
  vals <-  rbind(vals, output)
}


colnames(vals) <-c("Cultivar", "District", "RMSE_mean")
vals$RMSE_mean <- as.numeric(vals$RMSE_mean)

knitr::kable(vals, caption = "Cultivar assessement based on ensemble RMSE of average yields from RHEAS ensemble.")

knitr::kable(sort(unique(vals$Cultivar)), caption = "Revelevant cultivars for Malawi")
```


Visualize RMSE spatial distribution.


```{r v3, echo=FALSE, warning=FALSE, message=FALSE}
library(raster)
filename <- "D:/Adm data/Malawi/gadm40_MWI_1.shp"
mwi <- shapefile(filename)
names(mwi)[4] <- "District"
mwi$District <- toupper(mwi$District)
mwi <-  merge(mwi[,"District"], vals, by = "District")
library(tmap)
library(mapview)
tmap_mode("view")
 map <- tm_shape(mwi, name="Prediction Error MT/ha") +
  tm_fill("RMSE_mean", title="RMSE", breaks = c(0.1,0.4,0.8,1.2,1.6,1.8), textNA = "No data") +
   tm_layout(title.size =5, panel.label.size=6)+
   tm_format("World")
 map 
```


Cultivars 4, 6 and 7 seem to show lower RMSE on mean of RHEAS ensemble yield forecasts. These cultivars corresponds to the following names. 

| No. | Cultivar Name |
| --- | ------------- |
| 1   | SC627         |
| 2   | SC403         |
| 3   | PAN53         |
| 4   | DKC8053       |
| 5   | PHB30G1       |
| 6   | DKC8033       |
| 7   | SC627         |
| 8   | ZMS 606       |
| 9   | PHB 30G19     |
| 10  | PHB 30B50     |
| 11  | SC513         |
| 12  | MRI624        |


