---
title: "Malawi Maize Yield Forecast"
author: "Dr. -Ing. MISK Benson Kenduiywo"
csl: "apa-6th-edition.csl"
output: 
   html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60))
opts_knit$set(global.par = TRUE)
```

## Reference data

Load Zambia's ministry of agriculture maize forecasts data as reference.

```{r ref1, echo=FALSE, warning=FALSE, message=FALSE}
rm(list = ls(all=TRUE))
unlink(".RData")
root <- "D:/RCMRD/Data/Yields/"
r.file <- paste0(root, "Reference/Malawi/MOA/MALAWI_MOA_MAIZE_Production data.csv")
```

Load data

```{r ref40, echo=FALSE, warning=FALSE, message=FALSE}
ref <- read.csv(r.file, stringsAsFactors =  FALSE)
ref$District <- toupper(ref$District)
ref$yield_MT_ha <- as.numeric(ref$yield_MT_ha)
ref <- aggregate(yield_MT_ha~District+Year, data=ref, FUN=mean, na.rm=T)
```


## RHEAS forecasts

Load and aggregate RHEAS simulated Leaf Area Index (LAI), Water stress and Grain Weight Average Dry (GWAD) across different ensembles. Extract year from dates (we will use harvest year).

```{r rh1, echo=FALSE, message=F}
tt <- read.csv(paste0(root, "RHEAS/malawi_chirps_dssatTable_2001_2022.csv"), stringsAsFactors =  FALSE)
tt$harvest <- as.Date(tt$harvest)
tt$planting <- as.Date(tt$planting)
tt$date <- format(tt$harvest, format = "%Y")
names(tt)[3] <- "District"
tt$District <- toupper(tt$District)
```


## Metrics aggregagation

Aggregate RHEAS production forecasts and metrics with respect to Districts maize growing calendar. 

There is one maize growing seasons in Malawi than runs from October to May/June.

So we will aggregate the metrics and yield forecast per district with this condition using the function `RH_metrics`.

```{r rh2, echo=F, message=F}
RH_metrics <- function(rh, sStart, sEnd, season){
  rh$date <- format(as.Date(tt$planting),  "%Y")
  rh$Season <- season
  rh <- aggregate(rh[,c("wsgd","lai","gwad")], rh[,c("District","date", "Season")], mean, na.rm=T)
  return(rh)
}

rh <- RH_metrics(tt, sStart ="10", sEnd = "05", "1")

```

Convert RHEAS yields from kg/ha to MT/ha.

```{r rh3, echo=FALSE}
rh$gwad <- rh$gwad/1000
```


## Visualization

Add shapefile for visualization.

```{r, v1, echo=F, message=F}
library(raster)
filename <- "D:/Adm data/Malawi/gadm40_MWI_1.shp"
mwi <- shapefile(filename)
names(mwi)[4] <- "District"
mwi$District <- toupper(mwi$District)
```


Check and format District names to be consistent in both the RHEAS and administrative boundaries.

```{r v2, echo=F, message=F}
c <- sort(unique(mwi$District))
c[!c %in% sort(unique(rh$District))]
```

 Merge RHEAS and Admin data.
 
```{r v3, echo=F, message=F}
mwi <-  merge(mwi[,"District"], rh, by = "District", duplicateGeoms=TRUE)
path <- "D:/RCMRD/Code/yieldprediction"
shapefile(mwi[,c("District", "wsgd", "lai","gwad", "date","Season")], paste0(path,"/malawi/malawi_Maize_Forecasts.shp"),overwrite=T)

```


Visualize RHEAS predicted yields spatially. 

```{r v4, echo=F, message=F}
library(tmap)
library(mapview)
mwi.p  <- mwi
year <- 2021
mwi.p <- subset(mwi.p, date==year)
tmap_mode("view")
map_a <- tm_shape(mwi.p, name="Average yield (MT/ha) per District") +
  tm_polygons(col=c("gwad", "lai"),  n=5, title=c(paste(year, " Yield (MT/ha)"), "Leaf Area Index")) +
  tm_facets(sync = TRUE, ncol = 2, nrow=1) +
  #tm_text("District", size = 0.75)+
  tm_layout(title.size =5, panel.label.size=6)+
  tm_format("World")
map_a

```


## Annual Seasonal Trends

Visualize trends for the last 5 years.
 
```{r bb1, echo=FALSE}
library(ggplot2)
temp <- rh[rh$date > 2009,]
p <- ggplot(temp, aes(x = date, y = gwad))

p <- p + geom_boxplot(
  aes(fill = Season),
  position = position_dodge(0.9) 
  ) +
  scale_fill_manual(values = c("#999999", "#E69F00"))+
  ggtitle("Annual Seasonal  Maize Yield")+
  labs(y = "Yield (MT/ha)", x = "Year")
p

```
 
 
## Validation

Compare RHEAS with MoA forecasts.

Merge the MoA forecasts with RHEAS ones by year and District.

```{r val2, echo=FALSE, warning=FALSE, message=FALSE}
x <- rh
names(x)[2] <- "Year"
y <- ref
dff <-  merge(x, y, by = c("District", "Year"))
dff <- na.omit(dff)
```


### RMSE

We can use the Root Mean Square Error (RMSE)  and mean absolute percentage error (MAPE) to evaluate the models accuracy. RMSE is given as:

$$
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i=1}^n y-\widehat{y}},
$$
where $\widehat{y}$  and $y$ are predicted yields and observed yields respectively while *n* is the number of fitted points.

```{r rmse, echo=FALSE, warning=FALSE, message=FALSE}
rmse <- function(error){
  sqrt(mean(error^2))
}

```

Compute RMSE.

```{r val5, echo=FALSE, warning=FALSE, message=FALSE}

cat("RMSE ", rmse(dff$gwad-dff$yield_MT_ha), "\n")

```

### RRMSE

What about Relative root mean square error (RRMSE):

$$
\text{RRMSE} = \frac{\sqrt{\frac{1}{n} \sum_{i=1}^n y-\widehat{y}}}{\frac{1}{n} \sum_{i=1}^n y},
$$


```{r rrmse, echo=FALSE, warning=FALSE, message=FALSE}
rrmse <- function(predicted, observed){
  error <- observed - predicted
  re <- sqrt(mean(error^2))/mean(observed)
  return(re*100)
}
```


Compute RRMSE.

```{r val6, echo=FALSE, warning=FALSE, message=FALSE}

cat("RRMSE ", rrmse(dff$gwad, dff$yield_MT_ha), "\n")

```

According to [Li et al. 2013](ttp://dx.doi.org/10.1016/j.enconman.2013.03.004), the performance of the model is excellent when RRMSE < 10%; good if 10\% < RRMSE < 20\%; fair if 20\% < RRMSE < 30\%; and poor if RRMSE ≥ 30\%.


 