---
title: "Kenya MAIZE Forescasts"
author: "Dr. -Ing. MISK Benson Kenduiywo"
csl: "apa-6th-edition.csl"
output: 
   html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r, global_options, tidy = TRUE,echo=TRUE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60))
opts_knit$set(global.par = TRUE)
```

## Reference data

Load the reference maize yield data from [MOALF](http://kilimodata.developlocal.org/). First define paths.

```{r ref1, echo=F, message=F, warning=F}
rm(list = ls(all=TRUE))
unlink(".RData")
root <- "D:/RCMRD/Data/Yields/"

```

Load data

```{r ref40, echo=F, message=F, warning=F}
#Load 2010-2018 data
r1 <- read.csv(paste0(root, "Reference/Maize production by County MOALF 2010-2017.csv"), stringsAsFactors =  FALSE)
ha <- r1[r1$unit=="ha", c("county", "year", "amount")]
colnames(ha)[3] <- "area_ha"
mt <- r1[r1$unit=="mt", c("county", "year", "amount")]
colnames(mt)[3] <- "Yield_MT"
r1 <- merge(mt, ha, by = c("county", "year"))
colnames(r1)[1] <- "County"
colnames(r1)[2] <- "date"
r1$Yield_MT_ha <- r1$Yield_MT/r1$area_ha

#Add year 2018
temp <- read.csv(paste0(root, "Reference/Maize production by County MOALF 2018.csv"), stringsAsFactors =  FALSE)
#agg = aggregate(.~County+date, data=temp, mean, na.rm=T)
agg <- aggregate(temp[,c("Yield_MT","area_ha"), drop=FALSE], temp[, c("County","date"), drop=FALSE], sum, na.rm=TRUE)
agg$Yield_MT_ha <- agg$Yield_MT/agg$area_ha

r1$County <- toupper(r1$County)
ref <- rbind(r1,agg)
```

Check number of counties

```{r ref5, echo=F, message=F, warning=F}
kable(head(sort(unique(ref$County))), caption="Counties")
```

Visualize the maize production estimates from data from MoA per year.

```{r par, echo=F, message=F, warning=F}
par(mar=c(2,2,2,2)) #c(bottom, left, top, right)
```


```{r ref3, echo=F, message=F, warning=F}
names(ref)[2] <- "Year"
boxplot(Yield_MT_ha~Year, data=ref, col=rainbow(length(unique(ref$Year))), xlab="Year", ylab = "Yield (MT/ha)", main="MoALF Annual Forecasts.")

```

Visualize the crop forecasting data from MoA per County

```{r ref4, echo=F, message=F, warning=F}
agg <- ref
agg$County <- substr(agg$County,start=1,stop=3)
agg <- na.omit(agg)
boxplot(Yield_MT_ha~County, data=agg, col=rainbow(length(unique(agg$Province))), xlab="County", ylab = "Yield (MT/ha)", main="Zambia MoA Provinces Forecasts.", cex.axis=0.8) #horizontal = T, las=1

```

Aggregate MoALF forecasts per County

```{r ref10, echo=F, message=F, warning=F}
ref.agg <- aggregate(ref[,"Yield_MT_ha", drop=FALSE], ref[, c("County","Year"), drop=FALSE], mean, na.rm=TRUE)
```


## RHEAS simulated yields

Load and aggregate RHEAS simulated Leaf Area Index (LAI), Water stress and Grain Weight Average Dry (GWAD) across different ensembles. Extract year from dates (we will use harvest year).

```{r rh1}
tt <- read.csv(paste0(root, "RHEAS/kenya_dssatTable_1990_2022.csv"), stringsAsFactors =  FALSE)
tt$harvest <- as.Date(tt$harvest)
tt$planting <- as.Date(tt$planting)
tt$date <- format(tt$harvest, format = "%Y")
names(tt)[3] <- "County"
```


## Production forecast and RHEAS metrics aggregagation

Aggregate RHEAS production forecasts and metrics with respect to County maize growing calendar. 

Generally the maize growing season in Kenya can be grouped into two e.g. March--October and September--February. So we will aggregate the metrics and forecast with this condition using the function `RH_metrics`.

```{r rh2}
RH_metrics <- function(rh, sStart, sEnd, season){
  rh <- subset(rh, format(as.Date(rh$planting), "%m") >= sStart & format(as.Date(rh$harvest), "%m") <= sEnd)
  rh$Season <- season
  rh <- aggregate(rh[,c("wsgd","lai","gwad")], rh[,c("County","date", "Season")], mean, na.rm=T)
  return(rh)
}

rh1 <- RH_metrics(tt, sStart ="09", sEnd = "02", "1")
rh2 <- RH_metrics(tt, sStart ="03", sEnd = "11", "2")
rh <- rbind(rh1,rh2)
rh$County <- toupper(rh$County)
```

Convert RHEAS yields from kg/ha to MT/ha.

```{r rh3}
rh$gwad <- rh$gwad/1000
```


## Visualization

Add shapefile for visualization.

```{r, v1}
library(raster)
filename <- "D:/Adm data/Kenya_counties_2011/Kenya_county_dd.shp"
ke <- shapefile(filename)
names(ke)[1] <- "County"
ke$County <- toupper(ke$County)
```


Check and format County names to be consistent in both the RHEAS forecasts and that of administrative boundaries.

```{r v2}
#RHEAS
c <- sort(unique(rh$County))
c[!c %in% sort(unique(ke$County))]
#MoA Forecasts
c <- sort(unique(ref$County))
c[!c %in% sort(unique(ke$County))]
```

 Merge RHEAS and Admin data.
 
```{r v3}
ke <-  merge(ke[,"County"], rh[rh$date=="2022",], by = "County", duplicateGeoms=TRUE)
path <- "D:/RCMRD/Code/yieldprediction"
shapefile(ke[,c("County", "wsgd", "lai","gwad")], paste0(path,"/kenya/kenya_Maize_Forecasts.shp"), overwrite=T)

```

Visualize RHEAS predicted yields spatially.

### First Season

Visualize RHEAS predicted yields spatially for *season 1*. *NOTE: Ignore yields over the lake area*.

```{r v4, echo=F, message=F}
library(tmap)
library(mapview)
ke.p  <- ke
year <- 2022
ke.p <- subset(ke.p, date==year & Season==1)
tmap_mode("view")
map_a <- tm_shape(ke.p, name="Average yield (MT/ha) per County") +
  tm_polygons(col=c("gwad", "lai"), title=c(paste(year, " Yield (MT/ha)"), "Leaf Area Index")) +
  tm_facets(sync = TRUE, ncol = 2, nrow=1) +
  #tm_text("District", size = 0.75)+
  tm_layout(title.size =5, panel.label.size=6)+
  tm_format("World")
map_a

```

The first season seems to do well on average.

### Second Season

Now lets how  *season 2* looks spatially

```{r v5, echo=F , message=F}
library(tmap)
library(mapview)
ke.p  <- ke
year <- 2022
ke.p <- subset(ke.p, date==year & Season==2)
tmap_mode("view")
map_a <- tm_shape(ke.p, name="Average yield (MT/ha) per County") +
  tm_polygons(col=c("gwad", "lai"), title=c(paste(year, " Yield (MT/ha)"), "Leaf Area Index")) +
  tm_facets(sync = TRUE, ncol = 2, nrow=1) +
  #tm_text("District", size = 0.75)+
  tm_layout(title.size =5, panel.label.size=6)+
  tm_format("World")
map_a

```


## Validation

Compare RHEAS with MoA forecasts.

First check consistency in Districts names.

```{r val1}
c <- sort(unique(rh$County))
c[!c %in% sort(unique(ref.agg$County))]
```

Merge the MoA forecasts with RHEAS ones by year and District.

```{r val2}
baseYear <- 2010
x <- rh[rh$date>=baseYear,]
names(x)[2] <- "Year"
y <- ref.agg[ref.agg$Year>=baseYear,]
dff <-  merge(x,y, by = c("County", "Year"))

```

Plot RHEAS forecast versus MoA forecasts.

```{r val3}
#dff <- dff[dff$Yield_MT_ha <= 4.5,]
x11()
plot(gwad~Yield_MT_ha, data=dff, pch=16, xlim = c(0,4.5), ylim = c(0,4.5), xlab="MoA Forecasts (MT/ha)", ylab="RHEAS Forecasts (MT/ha)")
abline(0,1, lwd=2, col="red")
```


Plot the two datasets trends.

```{r val4}
dff.agg <- aggregate(dff[,c("Yield_MT_ha", "gwad"), drop=FALSE], dff[, "Year", drop=FALSE], mean, na.rm=TRUE)

x11()
plot(dff.agg$Year, dff.agg$Yield_MT_ha, type="l", lwd=2, col="red", ylim = c(0,4), ylab="Maize Forecasts (MT/ha)", xlab="Year")
lines(dff.agg$Year, dff.agg$gwad, col="blue", lwd=2)
legend("topright", c("MoA", "RHEAS"), col=c("red","blue"), lty = 1, lwd=2)

```



```{r test}
par(mfrow=c(1,2))
dff$County <- substr(dff$County, start=1,stop=3)
boxplot(Yield_MT_ha~Year, data=dff, col=rainbow(length(unique(ref$Year))), xlab="Year", ylab = "Yield (MT/ha)", main="MoA Annual Forecasts.")

boxplot(gwad~Year, data=dff, col=rainbow(length(unique(ref$Year))), xlab="Year", ylab = "Yield (MT/ha)", ylim=c(0,4.5), main="RHEAS Annual Forecasts.")
```

We can use the Root Mean Square Error (RMSE)  and mean absolute percentage error (MAPE) to evaluate the models accuracy. RMSE is given as:

$$
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i=1}^n \widehat{y}-y},
$$
where $\widehat{y}$  and $y$ are predicted yields and observed yields respectively while *n* is the number of fitted points.

```{r rmse}
rmse <- function(error){
  sqrt(mean(error^2))
}

```

Compute RMSE.

```{r val5}
rmse(dff$gwad-dff$Yield_MT_ha)
```


