---
title: "Zambia MAIZE Forescasts"
author: "Dr. -Ing. MISK Benson Kenduiywo"
csl: "apa-6th-edition.csl"
output: 
   html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r, global_options, tidy = TRUE,echo=TRUE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60))
opts_knit$set(global.par = TRUE)
```

## Reference data

Load Zambia's ministry of agriculture maize forecasts data as reference.

```{r ref1, echo=FALSE, warning=FALSE, message=FALSE}
rm(list = ls(all=TRUE))
unlink(".RData")
root <- "D:/RCMRD/Data/Yields/"
r.file <- paste0(root, "Reference/Zambia/MOA/Zambia_District _Maize_Forecasting_2011_2022.csv")
```


Load data

```{r ref40, echo=FALSE, warning=FALSE, message=FALSE}
ref <- read.csv(r.file, stringsAsFactors =  FALSE)
names(ref)
```

Convert District and province names to upper case.

```{r ref5, echo=FALSE, warning=FALSE, message=FALSE}
ref$Province <- trimws(toupper(ref$Province), which = c("both"))
sort(unique(ref$Province))
ref$District <- trimws(toupper(ref$District), which = c("both"))
ref$District <- toupper(ref$District)
ref$District[ref$District=="ITEZHI-TEZHI"] <- "ITEZHI TEZHI"
kable(head(sort(unique(ref$District))), caption="Districts")
```

Visualize the crop forecasting data from MoA per year.

```{r par, echo=FALSE, warning=FALSE, message=FALSE}
par(mar=c(2,2,2,2)) #c(bottom, left, top, right)
```


```{r ref3}

boxplot(yield_MT_ha~Year, data=ref, col=rainbow(length(unique(ref$Year))), xlab="Year", ylab = "Yield (MT/ha)", main="Zambia MoA Annual Forecasts.")

```

Visualize the crop forecasting data from MoA per Province.

```{r ref4, echo=FALSE, warning=FALSE, message=FALSE}

agg <- ref#aggregate(ref[,"yield_MT_ha", drop=FALSE], ref[, c("Province","District","Year"), drop=FALSE], sum, na.rm=TRUE)
agg$Province <- substr(agg$Province,start=1,stop=4)
agg <- na.omit(agg)
boxplot(yield_MT_ha~Province, data=agg, col=rainbow(length(unique(agg$Province))), xlab="Provinces", ylab = "Yield (MT/ha)", main="Zambia MoA Provinces Forecasts.", cex.axis=0.8) #horizontal = T, las=1

```


Aggregate MoA forecasts per district.

```{r ref10, echo=FALSE, warning=FALSE, message=FALSE}
ref.agg <- aggregate(ref[,"yield_MT_ha", drop=FALSE], ref[, c("District","Year"), drop=FALSE], mean, na.rm=TRUE)
```


## RHEAS simulated yields

Load and aggregate RHEAS simulated Leaf Area Index (LAI), Water stress and Grain Weight Average Dry (GWAD) across different ensembles. Extract year from dates (we will use harvest year).

```{r rh1, echo=FALSE, warning=FALSE, message=FALSE}
tt <- read.csv(paste0(root, "RHEAS/zambia_tamsat_25km_districts_dssatTable_2000_2022_50kg.csv"), stringsAsFactors =  FALSE)
tt$harvest <- as.Date(tt$harvest)
tt$planting <- as.Date(tt$planting)
tt$date <- format(tt$harvest, format = "%Y")
names(tt)[3] <- "District"
tt$District[tt$District=="CHIENGI"] <- "CHIENGE"
tt$District[tt$District=="SHANGOMBO"] <- "SHANG'OMBO"
```


## Unit Production forecast and RHEAS metrics aggregagation

Aggregate RHEAS production forecasts and metrics with respect to Districts maize growing calendar. 

The maize growing season in Zambia starts from October to end of June. So we will aggregate the metrics and forecast with this condition using the function `RH_metrics`.

```{r rh2, echo=FALSE, warning=FALSE, message=FALSE}
RH_metrics <- function(rh, sStart, sEnd){
  rh <- subset(rh, format(as.Date(rh$planting), "%m") >= sStart & format(as.Date(rh$harvest), "%m") <= sEnd)
  rh <- aggregate(rh[,c("wsgd","lai","gwad")], rh[,c("District","date")], mean, na.rm=T)
  return(rh)
}

rh <- RH_metrics(tt, sStart ="10", sEnd = "06")
rh$District <- toupper(rh$District)
```

Convert RHEAS yields from kg/ha to MT/ha.

```{r rh3, echo=FALSE, warning=FALSE, message=FALSE}
rh$gwad <- rh$gwad/1000

```



## Validation

Compare RHEAS with MoA forecasts.

Merge the MoA forecasts with RHEAS ones by year and District.

```{r val2, echo=FALSE, warning=FALSE, message=FALSE}
x <- rh
names(x)[2] <- "Year"
y <- ref

dff <-  merge(x, y, by = c("District", "Year"))

#cat("Pearson correlation between RHEAS and MOA yields ", cor(dff$wsgd, dff$yield_MT_ha, method = "pearson"))

```


### Validation

### RMSE

We can use the Root Mean Square Error (RMSE)  and mean absolute percentage error (MAPE) to evaluate the models accuracy. RMSE is given as:

$$
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i=1}^n y-\widehat{y}},
$$
where $\widehat{y}$  and $y$ are predicted yields and observed yields respectively while *n* is the number of fitted points.

```{r rmse, echo=FALSE, warning=FALSE, message=FALSE}
rmse <- function(error){
  sqrt(mean(error^2))
}

```

Compute RMSE.

```{r val5, echo=FALSE, warning=FALSE, message=FALSE}

cat("RMSE ", rmse(dff$gwad-dff$yield_MT_ha), "\n")

```

### RRMSE

What about Relative root mean square error (RRMSE):

$$
\text{RRMSE} = \frac{\sqrt{\frac{1}{n} \sum_{i=1}^n y-\widehat{y}}}{\frac{1}{n} \sum_{i=1}^n y},
$$


```{r rrmse, echo=FALSE, warning=FALSE, message=FALSE}
rrmse <- function(predicted, observed){
  error <- observed - predicted
  re <- sqrt(mean(error^2))/mean(observed)
  return(re*100)
}
```


Compute RRMSE.

```{r val6, echo=FALSE, warning=FALSE, message=FALSE}

cat("RRMSE ", rrmse(dff$gwad, dff$yield_MT_ha), "\n")

```

According to [Li et al. 2013](ttp://dx.doi.org/10.1016/j.enconman.2013.03.004), the performance of the model is excellent when RRMSE < 10%; good if 10\% < RRMSE < 20\%; fair if 20\% < RRMSE < 30\%; and poor if RRMSE ≥ 30\%.


## Total production

Total production _P_ is estimated as a product of unit production _u_ (in MT/ha) and area under maize _a_ per district, that is,
$$

p = u \times a.
$$

```{r p1, echo=FALSE, warning=FALSE, message=FALSE}
library(readxl)
a <- read_excel("D:/RCMRD/Data/Maps/Zambia/ZMB_Maize_Area_Stats_s1_s2.xls", col_names = TRUE)
names(a)[2] <- "District"
a$District <- toupper(a$District)
cel_area <- (10*10)/10000 #Sentinel spatial resolution
a$Area_ha <- a$COUNT * cel_area 

```

Merge maize area statistics per district with unit production forecast and compute production in MT/ha.

```{r p2}
rh <- merge(rh, a[,c("District","Area_ha")], by="District")
rh$Production <- rh$gwad * rh$Area_ha
```


## Visualization

Add shapefile for visualization.

```{r, v1, echo=FALSE, warning=FALSE, message=FALSE}
library(raster)
filename <- "D:/Adm data/Zambia/2010 Districts/district_74.shp"
zmb <- shapefile(filename)
names(zmb)[3] <- "District"
zmb$District <- toupper(zmb$District)
zmb$PROVINCE <- toupper(zmb$PROVINCE)
```


Check and format District names to be consistent in both the RHEAS and administrative boundaries.

```{r v2, echo=FALSE, warning=FALSE, message=FALSE}
#RHEAS
c <- sort(unique(rh$District))
c[!c %in% sort(unique(zmb$District))]
#MoA Forecasts
c <- sort(unique(ref$District))
c[!c %in% sort(unique(zmb$District))]
```

Merge RHEAS and Admin data.
 
```{r v3, echo=FALSE, warning=FALSE, message=FALSE}
#zmb <-  merge(zmb[,"District"], rh[rh$date=="2022",], by = "District")
year <- 2022
zmb <-  merge(zmb[,c("District", "PROVINCE")], rh[rh$date=="2022",], by = "District")
path <- "D:/RCMRD/Code/yieldprediction"
shapefile(zmb[,c("District", "wsgd", "lai","gwad")], paste0(path,"/zambia/Zambia_Maize_Forecasts.shp"),overwrite=T)

boxplot(Production/1000~PROVINCE, data=zmb@data, col=rainbow(length(unique(zmb$PROVINCE))), xlab="Provinces", ylab = "Production in thousands of MT", main="2022 PRODUCTION FORECAST.", cex.axis=0.8) #horizontal = T, las=1

```


Visualize RHEAS predicted yields spatially.

```{r v4, echo=FALSE, warning=FALSE, message=FALSE}
library(tmap)
library(mapview)
tmap_mode("view") #breaks = seq(2,4.1,0.1)
map_a <- tm_shape(zmb, name="Average yield (MT/ha) per District") +
  tm_polygons(col=c("gwad", "Production"),  n=5, title=c(paste(year, " Yield (MT/ha)"), "Production (MT)"), textNA = "No data") +
  tm_text("District", size = 0.75)+
  tm_layout(title.size =5, panel.label.size=6)+
  tm_format("World")
map_a

#lf <- tmap_leaflet(map_a)
#mapshot(lf, file = "D:/RCMRD/Presentations/ISPRS_2022/Yields_2022_MODIS.png")

```

