---
title: "Zambia MAIZE Forescasts"
author: "Dr. -Ing. MISK Benson Kenduiywo"
csl: "apa-6th-edition.csl"
output: 
   html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r, global_options, tidy = TRUE, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60))
opts_knit$set(global.par = TRUE)
```

## Reference data

Zambia's Ministry of Agriculture (MoA) maize census survey reference data.

```{r ref1, echo=FALSE, warning=FALSE, message=FALSE}
rm(list = ls(all=TRUE))
unlink(".RData")
root <- "D:/RCMRD/Data/Yields/"
r.file <- paste0(root, "Reference/Zambia/MOA/Zambia_District _Maize_Forecasting_2011_2022.csv")
```


<!-- Load data -->

```{r ref40, echo=FALSE, warning=FALSE, message=FALSE}
ref <- read.csv(r.file, stringsAsFactors =  FALSE)

```

<!-- Convert District and province names to upper case. -->

```{r ref5, echo=FALSE, warning=FALSE, message=FALSE}
ref$Province <- trimws(toupper(ref$Province), which = c("both"))
ref$District <- trimws(toupper(ref$District), which = c("both"))
ref$District <- toupper(ref$District)
ref$District[ref$District=="ITEZHI-TEZHI"] <- "ITEZHI TEZHI"

```

<!-- Visualize the crop forecasting data from MoA per year. -->

```{r par, echo=FALSE, warning=FALSE, message=FALSE}
#par(mar=c(2,2,2,2)) #c(bottom, left, top, right)
```


```{r ref3, echo=FALSE, warning=FALSE, message=FALSE}

boxplot(yield_MT_ha~Year, data=ref, col=rainbow(length(unique(ref$Year))), xlab="Year", ylab = "Yield (MT/ha)", main="Zambia MoA Annual Forecasts.")

```

Zambia's Ministry of Agriculture (MoA) maize forecasts per Province.

```{r ref4, echo=FALSE, warning=FALSE, message=FALSE}

agg <- ref#aggregate(ref[,"yield_MT_ha", drop=FALSE], ref[, c("Province","District","Year"), drop=FALSE], sum, na.rm=TRUE)
agg$Province <- substr(agg$Province,start=1,stop=4)
agg <- na.omit(agg)
boxplot(yield_MT_ha~Province, data=agg, col=rainbow(length(unique(agg$Province))), xlab="Provinces", ylab = "Yield (MT/ha)", main="Zambia MoA Provinces Forecasts.", cex.axis=0.8) #horizontal = T, las=1

```


<!-- Aggregate MoA forecasts per district. -->

```{r ref10, echo=FALSE, warning=FALSE, message=FALSE}
ref.agg <- aggregate(ref[,"yield_MT_ha", drop=FALSE], ref[, c("District","Year"), drop=FALSE], mean, na.rm=TRUE)
```


## RHEAS Forecasts

<!-- Load and aggregate RHEAS simulated Leaf Area Index (LAI), Water stress and Grain Weight Average Dry (GWAD) across different ensembles. Extract year from dates (we will use harvest year). -->

```{r rh1, echo=FALSE, warning=FALSE, message=FALSE}
tt <- read.csv(paste0(root, "RHEAS/zambia_tamsat_25km_districts_dssatTable_2000_2022_50kg.csv"), stringsAsFactors =  FALSE)
tt$harvest <- as.Date(tt$harvest)
tt$planting <- as.Date(tt$planting)
tt$date <- format(tt$harvest, format = "%Y")
names(tt)[3] <- "District"
tt$District[tt$District=="CHIENGI"] <- "CHIENGE"
tt$District[tt$District=="SHANGOMBO"] <- "SHANG'OMBO"
```


<!-- ## Unit Production forecast and RHEAS metrics aggregagation -->

<!-- Aggregate RHEAS production forecasts and metrics with respect to Districts maize growing calendar.  -->

<!-- The maize growing season in Zambia starts from October to end of June. So we will aggregate the metrics and forecast with this condition using the function `RH_metrics`. -->

```{r rh2, echo=FALSE, warning=FALSE, message=FALSE}
RH_metrics <- function(rh, sStart, sEnd){
  rh <- subset(rh, format(as.Date(rh$planting), "%m") >= sStart & format(as.Date(rh$harvest), "%m") <= sEnd)
  rh <- aggregate(rh[,c("wsgd","lai","gwad")], rh[,c("District","date")], mean, na.rm=T)
  return(rh)
}

rh <- RH_metrics(tt, sStart ="10", sEnd = "06")
rh$District <- toupper(rh$District)
```

<!-- Convert RHEAS yields from kg/ha to MT/ha. -->

```{r rh3, echo=FALSE, warning=FALSE, message=FALSE}
rh$gwad <- rh$gwad/1000

```



### Total production

Total production _P_ is estimated as a product of unit production _u_ (in MT/ha) and area under maize _a_ per district, that is,

$$
p = u \times a.
$$

```{r p1, echo=FALSE, warning=FALSE, message=FALSE}
library(readxl)
a <- read_excel("D:/RCMRD/Data/Maps/Zambia/ZMB_Maize_Area_Stats_s1_s2.xls", col_names = TRUE)
names(a)[2] <- "District"
a$District <- toupper(a$District)
cel_area <- (10*10)/10000 #Sentinel spatial resolution
a$Area_ha <- a$COUNT * cel_area 

```

<!-- merge maize area statistics per district with unit production forecast and compute production in mt/ha. -->

```{r p2, echo=FALSE, warning=FALSE, message=FALSE}
rh <- merge(rh, a[,c("District","Area_ha")], by="District")
rh$Production <- rh$gwad * rh$Area_ha
```


<!-- ## Visualization -->

<!-- Add shapefile for visualization. -->

```{r, v1, echo=FALSE, warning=FALSE, message=FALSE}
library(raster)
filename <- "D:/Adm data/Zambia/2010 Districts/district_74.shp"
zmb <- shapefile(filename)
names(zmb)[3] <- "District"
zmb$District <- toupper(zmb$District)
zmb$PROVINCE <- toupper(zmb$PROVINCE)
```


<!-- Check and format District names to be consistent in both the RHEAS and administrative boundaries. -->

```{r v2, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#RHEAS
c <- sort(unique(rh$District))
c[!c %in% sort(unique(zmb$District))]
#MoA Forecasts
c <- sort(unique(ref$District))
c[!c %in% sort(unique(zmb$District))]
```

<!-- Merge RHEAS and Admin data. -->
 
```{r v3, echo=FALSE, warning=FALSE, message=FALSE}
#zmb <-  merge(zmb[,"District"], rh[rh$date=="2022",], by = "District")
year <- 2022
zmb <-  merge(zmb[,c("District", "PROVINCE")], rh[rh$date=="2022",], by = "District")
path <- "D:/RCMRD/Code/yieldprediction"
shapefile(zmb[,c("District", "wsgd", "lai","gwad")], paste0(path,"/zambia/Zambia_Maize_Forecasts.shp"),overwrite=T)

production <- aggregate(zmb@data[, c("Area_ha", "Production"), drop =FALSE], zmb@data[, "PROVINCE", drop =FALSE], sum, na.rm=T)
temp <-  aggregate(zmb@data[,  "gwad", drop =FALSE], zmb@data[, "PROVINCE", drop =FALSE], mean, na.rm=T)
temp$gwad <- round(temp$gwad, 2)

production <- merge(temp, production, by="PROVINCE")
  
names(production)[2] <- "UNIT PRODUCTION (MT/ha)"
names(production)[3] <- "AREA (ha)"
names(production)[4] <- "PRODUCTION (MT)"

knitr::kable(production, align = 'c', caption = "Zambia's 2022 provincial Maize forecasts: Unit production (MT/ha), area under maize (ha), and total production (MT).")

zmb$PROVINCE <- substr(zmb$PROVINCE, 1,4)
boxplot(Production/1000~PROVINCE, data=zmb@data, col=rainbow(length(unique(zmb$PROVINCE))), xlab="Provinces", ylab = "Production in thousands of MT", main="2022 PRODUCTION FORECAST.", cex.axis=0.8) #horizontal = T, las=1

```

The area under maize was estimated through remote sensing mapping.

<!-- Visualize RHEAS predicted yields spatially. -->

```{r v4, echo=FALSE, warning=FALSE, message=FALSE}
library(tmap)
library(plot)
tmap_mode("view") #breaks = seq(2,4.1,0.1)
map_a <- tm_shape(zmb, name="Maize Yield per District") +
  tm_polygons(col=c("gwad", "Production"),  n=5, title=c(paste(year, " Yield (MT/ha)"), "Production (MT)"), textNA = "No data") +
  tm_facets(sync = TRUE, ncol = 2)+
  tm_layout(title.size =5, panel.label.size=6)+
  tm_legend(position = c("left", "top"))+ 
  tm_format("World")
#map_a
tmap_mode("view")
aa=qtm(zmb, fill = c("gwad", "Production"), fill.title=c("Yield (MT/ha)", "Production (MT)"), textNA = "No data")+
  tm_facets(sync = TRUE, ncol = 2)

#lf <- tmap_leaflet(map_a)
#mapshot(lf, file = "D:/RCMRD/Code/ZMB_Yields_2022.png")
tmap_save(aa, scale =1.6, dpi= 300, filename="D:/RCMRD/Code/ZMB_Yields_2022.png")
```

