---
title: "Rwanda Maize Yield Forecast"
author: "Dr. -Ing. MISK Benson Kenduiywo"
csl: "apa-6th-edition.csl"
output: 
   html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60))
opts_knit$set(global.par = TRUE)
```

## RHEAS forecasts

Load and aggregate RHEAS simulated Leaf Area Index (LAI), Water stress and Grain Weight Average Dry (GWAD) across different ensembles. Extract year from dates (we will use harvest year).

```{r rh1, echo=FALSE, message=F}
rm(list = ls(all=TRUE))
unlink(".RData")
root <- "D:/RCMRD/Data/Yields/"
tt <- read.csv(paste0(root, "RHEAS/rwanda_dssatTable_1990_2022.csv"), stringsAsFactors =  FALSE)
tt$harvest <- as.Date(tt$harvest)
tt$planting <- as.Date(tt$planting)
tt$date <- format(tt$harvest, format = "%Y")
names(tt)[3] <- "District"
tt$District <- toupper(tt$District)
```


## Metrics aggregagation

Aggregate RHEAS production forecasts and metrics with respect to Districts maize growing calendar. 

There two maize growing seasons in Rwanda [(One Acre Fund)](https://oneacrefund.org/blog/one-acre-fund-rwanda-launches-joint-venture-produce-maize-seed/#:~:text=Rwanda%20traditionally%20has%20two%20growing,in%20the%20shorter%20Season%20B.):
1) Season 1: sowing in September and harvesting January.
2) Season 2: sowing in February and harvesting in June. 

So we will aggregate the metrics and yield forecast per district with this condition using the function `RH_metrics`.

```{r rh2, echo=F, message=F}
RH_metrics <- function(rh, sStart, sEnd, season){
  rh <- subset(rh, format(as.Date(rh$planting), "%m") >= sStart & format(as.Date(rh$harvest), "%m") <= sEnd)
  rh$Season <- season
  rh <- aggregate(rh[,c("wsgd","lai","gwad")], rh[,c("District","date", "Season")], mean, na.rm=T)
  return(rh)
}

rh1 <- RH_metrics(tt, sStart ="09", sEnd = "02", "1")
rh2 <- RH_metrics(tt, sStart ="01", sEnd = "08", "2")
rh <- rbind(rh1,rh2)

```

Convert RHEAS yields from kg/ha to MT/ha.

```{r rh3, echo=FALSE}
rh$gwad <- rh$gwad/1000
```


## Visualization

Add shapefile for visualization.

```{r, v1, echo=F, message=F}
library(raster)
filename <- "D:/Adm data/rwanda/gadm36_RWA_2.shp"
uga <- shapefile(filename)
names(uga)[7] <- "District"
uga$District <- toupper(uga$District)
```


Check and format District names to be consistent in both the RHEAS and administrative boundaries.

```{r v2, echo=F, message=F}
c <- sort(unique(uga$District))
c[!c %in% sort(unique(rh$District))]
```

 Merge RHEAS and Admin data.
 
```{r v3, echo=F, message=F}
uga <-  merge(uga[,"District"], rh, by = "District", duplicateGeoms=TRUE)
path <- "D:/RCMRD/Code/yieldprediction"
shapefile(uga[,c("District", "wsgd", "lai","gwad", "date","Season")], paste0(path,"/rwanda/rwanda_Maize_Forecasts.shp"),overwrite=T)

```


### First Season

Visualize RHEAS predicted yields spatially for *season 1*. *NOTE: Ignore yields over the lake area*.

```{r v4, echo=F, message=F}
library(tmap)
library(mapview)
uga.p  <- uga
year <- 2022
uga.p <- subset(uga.p, date==year & Season==1)
tmap_mode("view")
map_a <- tm_shape(uga.p, name="Average yield (MT/ha) per District") +
  tm_polygons(col=c("gwad", "lai"),  n=5, title=c(paste(2023, " Yield (MT/ha)"), "Leaf Area Index")) +
  tm_facets(sync = TRUE, ncol = 2, nrow=1) +
  #tm_text("District", size = 0.75)+
  tm_layout(title.size =5, panel.label.size=6)+
  tm_format("World")
map_a

```

The first season seems to do well on average.

### Second Season

Now lets how  *season 2* looks spatially

```{r v5, echo=F , message=F}
library(tmap)
library(mapview)
uga.p  <- uga
year <- 2022
uga.p <- subset(uga.p, date==year & Season==2)
tmap_mode("view")
map_a <- tm_shape(uga.p, name="Average yield (MT/ha) per District") +
  tm_polygons(col=c("gwad", "lai"),  n=5,  title=c(paste(year, " Yield (MT/ha)"), "Leaf Area Index"),n=5) +
  tm_facets(sync = TRUE, ncol = 2, nrow=1) +
  #tm_text("District", size = 0.75)+
  tm_layout(title.size =5, panel.label.size=6)+
  tm_format("World")
map_a

```


## Annual Seasonal Trends

Visualize trends for the last 5 years.
 
```{r bb1, echo=FALSE}
library(ggplot2)
temp <- rh[rh$date > 2009,]
p <- ggplot(temp, aes(x = date, y = gwad))

p <- p + geom_boxplot(
  aes(fill = Season),
  position = position_dodge(0.9) 
  ) +
  scale_fill_manual(values = c("#999999", "#E69F00"))+
  ggtitle("Annual Seasonal  Maize Yield")+
  labs(y = "Yield (MT/ha)", x = "Year")
p

```
 